---
title: "Exploratory Data Analysis: Maternity"
subtitle: "Donald Dinerman, [holder], [holder]"  
date: 'June 24th, 2022'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#1381B0",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF"
)
```






```{r, echo=F}
#read data
maternity_df = read.csv("C:/Undergraduate/R/Optum-Bridge-Summer-Research/EDA Project/maternity.csv")
```

```{r, echo=F}
library(kableExtra)
library(tidyverse)

#rename to shorter colnames
maternity_df = rename(maternity_df, BirthsDeceased = PriorBirthsNowDeceased, Diabetes = PrePregnancyDiabetes, Hypertension = PrePregnancyHypertension)

main_tab = maternity_df %>% 
  select(State:Births) %>% 
  slice(c(1, 55, 99, 175, 252)) %>%  
  kbl() %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center")
```

```{r, echo=F}
#Hypothesis 1

#create dataset with counterfactual
h1_df = filter(maternity_df, TobaccoUse %in% c("Yes", "No"), Diabetes == "No", Hypertension == "No",
               BirthsDeceased == "0")

#Calculate standard deviation in group using rep(AverageBirthWeight, births)
h1_df_no = filter(h1_df, TobaccoUse == "No")

no_hold = NULL

for(i in 1:length(h1_df_no)){
  
  temp = rep(h1_df_no[i,]$AverageBirthWeight, h1_df_no[i,]$Births)
  
  no_hold = c(no_hold, temp)  
}

h1_df_yes = filter(h1_df, TobaccoUse == "Yes")

yes_hold = NULL

for(i in 1:length(h1_df_yes)){
  
  temp = rep(h1_df_yes[i,]$AverageBirthWeight, h1_df_yes[i,]$Births)
  
  yes_hold = c(yes_hold, temp)
  
}

#Barplot with CI (The CI are so small bc the n is so large)

#Tobacco Use is Associated with Decreased Birthweights
h1_plot = h1_df %>%
  group_by(TobaccoUse) %>%
  summarise(true_count = sum(Births), 
            mean_AverageBirthWeight = mean(AverageBirthWeight)) %>%
  mutate(total = sum(true_count),
         sd = c(sd(no_hold), sd(yes_hold)),
         margin = qt(0.975, df = true_count - 1)*sd/sqrt(true_count),
         lower = mean_AverageBirthWeight - margin,
         upper = mean_AverageBirthWeight + margin) %>%
  ggplot(aes(x = TobaccoUse, y = mean_AverageBirthWeight)) +
  geom_col(col = "black", fill = "steelblue") +
  #geom_errorbar(aes(ymin = lower,ymax = upper), color = "red") + #CI
  coord_cartesian(ylim = c(0, 3600)) + #Zoom in
  geom_text(aes(label = round(mean_AverageBirthWeight, 0)), vjust = -0.25) + #text label
  theme_bw() + 
  labs(title = "", x = "Tobacco Use", y = "Average Birthweight (Grams)") 
```

```{r, echo=F}
#Hypothesis 2

#Older Mothers Have More Prenatal Visits
h2_plot = maternity_df %>%
  ggplot(aes(x = AverageMotherAge, y = AverageNumberPrenatalVisits)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = F, method = "lm") +
  theme_bw() +
  labs(x = "Mother Age (Year)", y = "Number of Prenatal Visits", title = "") 
```

```{r, echo=F}
df_count = filter(maternity_df, BirthsDeceased != "Unknown", TobaccoUse != "Unknown") %>%
  mutate(new_mortality = ifelse(BirthsDeceased == "0", "None", "At least One")) %>%
  group_by(new_mortality, TobaccoUse) %>%
  summarise(count = sum(Births))

pre_table_df = data.frame(mortality = c(rep("At Least One", df_count$count[1]), rep("At Least One", df_count$count[2]), 
                                        rep("None", df_count$count[3]), rep("None", df_count$count[4])),
              tobacco = c(rep("No", df_count$count[1]), rep("Yes", df_count$count[2]), 
                          rep("No", df_count$count[3]), rep("Yes", df_count$count[4])))
```

```{r, echo=F}
#An alternative: Using context of baby deceased # to infer k (e.g., k = 2 bc 2 groups of deceased births)
library(dslabs)

#remove NA from df
cluster_df = dplyr::select(maternity_df, BirthsDeceased, AveragePrePregnancyBMI, AverageBirthWeight) %>%  drop_na()

#std cont var
clean_cluster_df <- cluster_df %>%
  mutate(std_BMI = as.numeric(scale(AveragePrePregnancyBMI, center = TRUE, scale = TRUE)), #<<
         std_weight = as.numeric(scale(AverageBirthWeight, center = TRUE, scale = TRUE)),
         update_deceased = ifelse(BirthsDeceased == "0", "None", "At Least One")) #<<

#K-means++
library(flexclust)

set.seed(12)

init_kmeanspp <- 
  kcca(dplyr::select(clean_cluster_df, #<<
                     std_BMI, std_weight), k = 2, #<<
       control = list(initcent = "kmeanspp")) #<<

cluster_plot = clean_cluster_df %>%
  mutate(country_clusters = 
           as.factor(init_kmeanspp@cluster)) %>% #<< #@ symbol bc init_kmeanspp is an s4 object
  ggplot(aes(x = AveragePrePregnancyBMI, y = AverageBirthWeight,
             color = country_clusters)) +
  geom_point() + 
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(color = "Cluster", x = "Mother's BMI", y = "Birth Weight (Grams)")

real_plot = clean_cluster_df %>%
  ggplot(aes(x = AveragePrePregnancyBMI, y = AverageBirthWeight,
             color = update_deceased)) +
  geom_point() +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(color = "Deceased Births #", x = "Mother's BMI", y = "Birth Weight (Grams)")
```


## Background and Motivation





---


## The Maternal Health Care Disparities Dataset


Loaded in a dataset of available domestic birth records from 2019

 - There's 3.7M observations aggregated into 842 rows

 - Rows represent unique states of births deceased, tobacco use, diabetes, and hypertension combinations

 - Categorical Variables: State, BirthsDeceased, TobaccoUse, Diabetes, Hypertension

 - Continuous Variables: Births, AverageMotherAge, AverageBirthWeight, AverageBMI, AveragePrenatalVisits
 
```{r,echo=F, fig.align = 'center', fig.width = 8}
main_tab #better fit, less wide, bottom of slide (shorten var names on slide, and in code for better fit?)
```
 
---

## Hypotheses

1. Tobacco use is associated with reduced child birth weight among mother's without prior birth deaths.

2. As the age of the mother increases, the number of prenatal visits increases.

3. Tobacco use is more prevalent among mothers with at least 1 prior birth deceased.

---

## Tobacco Use is Associated with Decreased Birthweights


```{r, echo = F, fig.align = 'center'}
h1_plot
```

---


## Older Mothers Have More Prenatal Visits


```{r, echo = F, fig.align = 'center'}
h2_plot
```

---

## High Tobacco Use for Mothers With at Least 1 Prior Birth Deceased


```{r, echo = F, fig.align = 'center'}
#High Tobacco Use for Mothers With at Least 1 Prior Birth Deceased
mosaicplot(table(pre_table_df$mortality, pre_table_df$tobacco),
           shade = T,
           main = " ", ylab = "Tobacco Use") 
```



---
## Natural Clusters in the Data


```{r, echo = F}
real_plot
```





---


## Kmeans++ Captures Natural Clusters 


```{r, echo = FALSE, fig.show="hold", out.width="50%"}
real_plot #check if sizing is ok, can stack them vertically instead

cluster_plot
```

- Accuracy : 60%






---


### Conclusions

- Mother's with prior tobacco use birth babies approximately 213 grams lighter than mother's without tobacco use.

- As mother's age increases, the number of prenatal visits increase. This may happen because older mothers are more susceptible to pregnancy complications.

- Among mother's with at least 1 prior birth deceased, approximately 11% used tobacco while only 6% of mother's without prior birth deaths used tobacco products.

### Limitations

- We did not perform any statistical tests (e.g., t-test) to compare group averages, so our conclusions are purely based on visualizations.

- Our clustering doesn't weigh the births within each row and rather treats each row as a single observation.

**Feel free to ask any questions!**


