---
title: "0613-Lecture"
output: html_document
date: '2022-06-13'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Density Estimation

Created dataset of shot attempts by the Stephen Curry in 2021-2022 season using [`nbastatR`](http://asbcllc.com/nbastatR/)

```{r, warning = FALSE, message = FALSE}
library(tidyverse)

curry_shots <- 
  read_csv("http://www.stat.cmu.edu/cmsac/sure/2022/materials/data/sports/xy_examples/curry_2022_shots.csv")
head(curry_shots)
```

- each row / observation is a shot attempt by Curry in the 2021 season
- __Categorical__ / qualitative variables: `is_shot_made`, `fg_type`, `shot_zone`, `shot_type`
- __Continuous__ / quantitative variables: `shot_x`, `shot_y`, `shot_distance`

## Histograms

```{r}
#Histogram
fd_bw <- 2 * IQR(curry_shots$shot_distance) / length(curry_shots$shot_distance)^(1/3) #set binwidth

curry_shots %>%
  ggplot(aes(x = shot_distance)) +
  geom_histogram(col = "black", fill = "steelblue", binwidth = fd_bw) +
  theme_bw()
```

__Small__ `binwidth`  $\rightarrow$ _"undersmooth"_ / spiky

```{r}
curry_shots %>%
  ggplot(aes(x = shot_distance)) +
  geom_histogram(col = "black", fill = "steelblue", binwidth = 1) + #<<
  theme_bw()
```

__Large__ `binwidth`  $\rightarrow$ _"oversmooth"_ / flat

```{r}
curry_shots %>%
  ggplot(aes(x = shot_distance)) +
  geom_histogram(col = "black", fill = "steelblue", binwidth = 25) +
  theme_bw()
```

By default the bins are centered on the integers...

- left-closed, right-open intervals 
- starting at -0.5 to 0.5, 0.5 to 1.5, ...

__Specify center of one bin__ (e.g. 0.5)

- Reminder to use `closed = "left"`...

```{r shot-dist-hist-shift, fig.height=4}
curry_shots %>%
  ggplot(aes(x = shot_distance)) +
  geom_histogram(col = "black", fill = "steelblue", 
                 binwidth = 1, center = 0.5, #center is 0.5*binwidth
                 closed = "left") +
  theme_bw()
```

How do histograms relate to the PDF and CDF?

__Remember__: we use the __probability density function (PDF)__ to provide a __relative likelihood__

- PDF is the __derivative__ of the cumulative distribution function (CDF)

- Histograms approximate the PDF with bins, and __points are equally likely within a bin__

```{r}
curry_shots %>%
  ggplot(aes(x = shot_distance)) + 
  geom_histogram(col = "black", fill = "steelblue",
                 binwidth = 1, center = 0.5, 
                 closed = "left") +
  geom_rug(alpha = 0.3) +
  theme_bw() +
  labs(x = "Shot distance (in feet)",
       y = "Number of shot attempts")
```

```{r}
curry_shots %>%
  ggplot(aes(x = shot_distance)) + 
  stat_ecdf() +
  geom_rug(alpha = 0.3) +
  theme_bw() +
  labs(x = "Shot distance (in feet)",
       y = "Proportion of Curry shot attempts")
```

__What can say about the relative likelihood of data we have not observed?__

- we want __non-zero density__ between our observations, e.g., just beyond 20 feet

## Kernel density estimation (KDE)

__Goal__: estimate the PDF $f(x)$ for all possible values (assuming it is continuous / smooth)

$$
\text{Kernel density estimate: } \hat{f}(x) = \frac{1}{n} \sum_{i=1}^n \frac{1}{h} K_h(x - x_i)
$$

- $n =$ sample size, $x =$ new point to estimate $f(x)$ (does NOT have to be in dataset!)

- $h =$ __bandwidth__, analogous to histogram binwidth, ensures $\hat{f}(x)$ integrates to 1

- $x_i =$ $i$th observation in dataset

- $K_h(x - x_i)$ is the __Kernel__ function, creates __weight__ given distance of $i$th observation from new point 
  - as $|x - x_i| \rightarrow \infty$ then $K_h(x - x_i) \rightarrow 0$, i.e. further apart $i$th row is from $x$, smaller the weight
  
  - as __bandwidth__ $h \uparrow$ weights are more evenly spread out (as $h \downarrow$ more concentrated around $x$) 

  - typically use [__Gaussian__ / Normal](https://en.wikipedia.org/wiki/Normal_distribution) kernel: $\propto e^{-(x - x_i)^2 / 2h^2}$
  
  - each kernel is centered about each data point, then kernels are added up to create the KDE
  
  - Big h means wide (smooth) kernel, small h means tall (sharp) kernel
  
  - $K_h(x - x_i)$ is large when $x_i$ is close to $x$

What about the bandwidth? See [Chapter 14 for more...]**(https://www.stat.cmu.edu/~cshalizi/ADAfaEPoV/)**

  Use Gaussian reference rule (rule-of-thumb) $\approx 1.06 \cdot \sigma \cdot n^{-1/5}$, where $\sigma$ is the observed standard deviation

  Modify the bandwidth using the `adjust` argument - __value to multiply default bandwidth by__

```{r}
curry_shots %>%
  ggplot(aes(x = shot_distance)) + 
  geom_density(adjust = 0.5) + #default adjust is determined by the Gaussian rule of thumb
  geom_rug(alpha = 0.3) + 
  theme_bw()
```

```{r}
curry_shots %>%
  ggplot(aes(x = shot_distance)) + 
  geom_density(adjust = 2) +
  geom_rug(alpha = 0.3) + 
  theme_bw()
```

Use density curves and ECDFs together

```{r}
library(patchwork)

curry_shot_dens <- curry_shots %>%
  ggplot(aes(x = shot_distance)) + 
  geom_density() +
  geom_rug(alpha = 0.3) +
  theme_bw() +
  labs(x = "Shot distance (in feet)",
       y = "Number of shot attempts")

curry_shot_ecdf <- curry_shots %>%
  ggplot(aes(x = shot_distance)) + 
  stat_ecdf() +
  geom_rug(alpha = 0.3) +
  theme_bw() +
  labs(x = "Shot distance (in feet)",
       y = "Proportion of Curry shot attempts")

curry_shot_dens + curry_shot_ecdf 
```

```{r}
curry_shot_dens_made <- curry_shots %>%
  ggplot(aes(x = shot_distance, 
             color = is_shot_made)) + 
  geom_density() +
  geom_rug(alpha = 0.3) +
  theme_bw() +
  labs(x = "Shot distance (in feet)",
       y = "Number of shot attempts")

curry_shot_ecdf_made <- curry_shots %>%
  ggplot(aes(x = shot_distance,
             color = is_shot_made)) + 
  stat_ecdf() +
  geom_rug(alpha = 0.3) +
  theme_bw() +
  labs(x = "Shot distance (in feet)",
       y = "Proportion of Curry shot attempts")

curry_shot_dens_made + curry_shot_ecdf_made + plot_layout(guides = 'collect')
```

## Alternative to violins - ridge plots

Useful to display conditional distributions across many levels

```{r}
library(ggridges)

curry_shots %>%
  ggplot(aes(x = shot_distance,
             y = shot_type)) + #<< 
  geom_density_ridges(rel_min_height = 0.01) + #<<
  theme_bw()
```

What about for 2D? (two continuous variables)

We can visualize all of the shot locations: (`shot_x`, `shot_y`)

```{r}
curry_shots %>%
  # Modify the shot coordinates
  mutate(shot_x = -shot_x / 10,
         shot_y = shot_y / 10) %>%
  ggplot(aes(x = shot_x, y = shot_y)) +
  geom_point(alpha = 0.3) +
  theme_bw()
```

## Create contours of 2D kernel density estimate (KDE)

- We make 2D KDE __contour__ plots using [`geom_density2d()`](https://ggplot2.tidyverse.org/reference/geom_density_2d.html)

```{r}
curry_shots %>%
  # Modify the shot coordinates
  mutate(shot_x = -shot_x / 10,
         shot_y = shot_y / 10) %>%
  filter(shot_y <= 30) %>%
  ggplot(aes(x = shot_x, y = shot_y)) + 
  geom_point(alpha = 0.3) + 
  geom_density2d() +
  theme_bw() + 
  theme(legend.position = "bottom") +
  coord_fixed() 
```

- Extend KDE for joint density estimates in 2D (see [section 14.4.2 for details](https://www.stat.cmu.edu/~cshalizi/ADAfaEPoV/))

- `coord_fixed()` forced a fixed ratio

- Can use `adjust` to modify the multivariate bandwidth

```{r}
curry_shots %>%
  # Modify the shot coordinates
  mutate(shot_x = -shot_x / 10,
         shot_y = shot_y / 10) %>%
  # Remove the outlier shots:
  filter(shot_y <= 30) %>%
  ggplot(aes(x = shot_x, y = shot_y)) + 
  geom_point(alpha = 0.3) + 
  geom_density2d(adjust = 0.3) +
  theme_bw() + 
  theme(legend.position = "bottom") +
  coord_fixed() #makes one unit on y equal to one unit on x
```

## Heatmaps

- We make 2D KDE __heatmap__ plots using [`stat_density_2d()`](https://ggplot2.tidyverse.org/reference/geom_density_2d.html)
and the `..` or [`after_stat()`](https://ggplot2.tidyverse.org/reference/aes_eval.html) function

- Can use white to a color to visualize density in heatmaps

```{r}
curry_shots %>%
  mutate(shot_x = -shot_x / 10,
         shot_y = shot_y / 10) %>%
  filter(shot_y <= 30) %>%
  ggplot(aes(x = shot_x, y = shot_y)) + 
  stat_density2d(h = 0.5, bins = 60,
                 aes(fill = after_stat(level)),
                 geom = "polygon") +
  scale_fill_gradient(low = "darkblue", 
                      high = "darkorange") +
  theme_bw() + 
  theme(legend.position = "right") +
  coord_fixed()
```

## Best alternative? Hexagonal binning

- We make __hexagonal heatmap__ plots using [`geom_hex()`](https://ggplot2.tidyverse.org/reference/geom_hex.html)

- Need to have the [`hexbin`](https://cran.r-project.org/web/packages/hexbin/index.html) package installed

```{r shot-loc-points-hex, eval = FALSE}
curry_shots %>%
  mutate(shot_x = -shot_x / 10,
         shot_y = shot_y / 10) %>%
  filter(shot_y <= 30) %>%
  ggplot(aes(x = shot_x, y = shot_y)) + 
  geom_hex(binwidth = c(1, 1)) +
  scale_fill_gradient(low = "darkblue", 
                      high = "darkorange") + 
  theme_bw() + 
  theme(legend.position = "right") +
  coord_fixed()
```

- Can specify `binwidth` in both directions
- Avoids limitations from smoothing

## What about his shooting efficiency?

- Can compute a function of another variable inside hexagons with [`stat_summary_hex()`](https://ggplot2.tidyverse.org/reference/stat_summary_2d.html)

- Check out [BallR](https://github.com/toddwschneider/ballr) for code examples to make shot charts and drawing courts


.pull-left[
```{r shot-loc-hex-make, eval = FALSE}
curry_shots %>%
  mutate(shot_x = -shot_x / 10,
         shot_y = shot_y / 10) %>%
  filter(shot_y <= 30) %>%
  ggplot(aes(x = shot_x, y = shot_y, 
             z = is_shot_made,
             group = -1)) +
  stat_summary_hex(binwidth = c(2, 2),
                   color = "black", 
                   fun = mean) + 
  scale_fill_gradient(low = "darkblue", 
                      high = "darkorange") + 
  theme_bw() + 
  theme(legend.position = "right") +
  coord_fixed()
```























