---
title: "0622-Lecture"
output: html_document
date: '2022-06-22'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Supervised Learning: Model assessment vs selection

What is supervised learning? 

Goal: uncover associations between a set of predictor (aka independent / explanatory) variables / features and a single response (or dependent) variable

- Ex: MLB records the batting average (hits / at-bats) for every player in each season. Can we accurately predict player X's batting average in year $t+1$ using their batting average in year $t$? Can we uncover meaningful relationships between other measurements (which are predictors / features) and batting average in year $t+1$ (which is the response)?

- Ex: We are provided player tracking data from the NFL, all x,y coordinates of every player on the field at every tenth of the second. Can we predict how far a ball-carrier will go at any given moment they have the football?

## Examples of statistical learning methods / algorithms

You are probably already familiar with statistical learning - even if you did not know what the phrase meant before

Examples of statistical learning algorithms include:

- Generalized linear models (GLMs) and penalized versions (Lasso, elastic net)

- Smoothing splines, Generalized additive models (GAMs)

- Decision trees and its variants (e.g., random forest, boosting)

- Neural networks (e.g., convolutional neural networks)

Two main types of estimation given values for predictors:

- Regression models: estimate _average_ value of response

- Classification models: determine _the most likely_ class of a set of discrete response variable classes

## Which method should I use in my analysis?

Depends on your goal - the big picture: inference vs prediction

Let $Y$ be the response variable, and $X$ be the predictors, then the learned model will take the form:

$$
\hat{Y}=\hat{f}(X)
$$

- Care about the details of $\hat{f}(X)$? $\Rightarrow$ You want to perform statistical inference

- Fine with treating $\hat{f}(X)$ as a obscure/mystical machine? $\Rightarrow$ Your interest is prediction

Any algorithm can be used for prediction, however options are limited for inference

Active area of research on using more mystical models for statistical inference

### Model flexibility vs interpretability

Generally speaking: tradeoff between a model's _flexibility_ (i.e. how "wiggly" it is) and how interpretable it is

- Simpler the parametric form of the model $\Rightarrow$ the easier it is to interpret

- Hence why linear regression is popular in practice

```{r out.width='50%', fig.align='center'}
library(tidyverse)
library(magick)

image_read("http://www.stat.cmu.edu/~pfreeman/flexibility.png") %>% print()
```
















