---
title: "Predictive Modeling"
output: html_document
date: '2022-06-29'
---

# Data Documentation
```{r}
#(https://www.countyhealthrankings.org/explore-health-rankings/measures-data-sources/county-health-rankings-model?componentType=factor-area&componentId=3)
```

Note: Data is aggregated and therefore variables must be weighted by population when feasible. By default, many variables are weighted by population.

**Response**

- Premature Death: Years of potential life (YPLL) lost before age 75 per 1000,000 population (2018-2020) 

**Predictors**
  
*Alcohol* 

- Excessive Drinking: Percentage of adults reporting binge or heavy drinking per 100,000 population (2019)
  
- Alcohol-Impaired Driving Deaths: Percentage of driving deaths with alcohol involvement (2016-2020)

*Diet and Exercise*

- Adult Obesity: Percentage of the adult population that reports a body mass index (BMI) greater than or equal to 30 kg/m2 (2019)
  
- Food Environment Index: Index of factors that contribute to a healthy food environment, from 0 (worst) to 10 (best)  (2019)
  
- Physical Inactivity: Percentage of adults age 18 and over reporting no leisure-time physical activity (age-adjusted) (2019)
  
- Access to Exercise Opportunities: Percentage of population with adequate access to locations for physical activity (2010 & 2021)
  
- Food Insecurity: Percentage of population who lack adequate access to food (2019)
  
- Limited Access to Healthy Foods: Percentage of population who are low-income and do not live close to a grocery store (2019)
  
*Tobacco Use*

- Adult Smoking: Percentage of adults (age 18 and older) who are current smokers (2019) 

*Sexual Activity* 

- Sexually Transmitted Infections (STI): Number of newly diagnosed chlamydia cases per 100,000 population (2019)

*Population*

- County Class: County description categorized by population (Urban $\geq$ 200k, Rural $\leq$ 50k, 200k $<$ Suburban $<$ 50k) (2020)

# Data Wrangling

```{r}
library(tidyverse)
library(readxl)
```

```{r, message=F, warning=F}
#r project establishes root wd so we can call files from the wd w/o explictly writing out the wd
data = read_xlsx("Data/CountyHealthRankingsData2022.xlsx", 4, skip = 1)

add_data = read_xlsx("Data/CountyHealthRankingsData2022.xlsx", 6, skip = 1)
```

```{r}
#remove features
#(: remove demographic info, CI: remove conf int, Quartile: remove quartile rankings)
#Note: May want to consider demographic features

new_data = select(data, -contains(c("CI", "Quartile"), ignore.case = F))

new_add_data = select(add_data, -contains(c("CI", "Quartile"), ignore.case = F))

comb_data = inner_join(new_data, new_add_data, by = c("FIPS", "State", "County"))
```

```{r}
#note: first row of a state is aggregated/averaged information at the state level
#note: Other rows are simply aggregated at the county level

#Identify and remove rows with unreliable data (Issue: This may affect counts for state level data)

reliable_data = comb_data %>%
  mutate(na_id = as.numeric(is.na(comb_data$Unreliable...4))) %>% #binary to is.na
  filter(na_id == 1) #keep only rows w NA (reliable rows)
```

```{r}
#Can Cluster by county attribute (on population lvl) (Population)
#County clustering (urban, suburban, rural) within state
#Capture county characteristics and state characteristics

#(https://www.hudexchange.info/faqs/programs/cdbg-entitlement-program/urban-county/where-is-the-term-urban-county-defined-within-the-cdbg-program-and-how/#:~:text=This%20provision%20defines%20an%20urban,including%20metropolitan%20cities%20located%20therein)
#(https://www.hhs.gov/guidance/document/defining-rural-population#:~:text=All%20counties%20that%20are%20not,as%20either%20Metro%20or%20Micro)
sub_upper_lim = 200000
sub_lower_lim = 50000

reliable_data = reliable_data %>%
  mutate(county_class = 
           case_when(Population < (sub_upper_lim) & Population > (sub_lower_lim) ~ "Suburban",
                     Population >= (sub_upper_lim) ~ "Urban", 
                     TRUE ~ "Rural")) #%>% select(State, County, county_class, Population)
```

# EDA

## Response Map
```{r}
library(usmap)
library(ggplot2)

#State
state_df = reliable_data %>%
  group_by(State) %>%
  dplyr::slice(1) %>%
  rename(., state = State, YPLL = 'Years of Potential Life Lost Rate', Life_Expectancy = 'Life Expectancy')

#YPLL (low values good)
state_map = plot_usmap(data = state_df, values = "YPLL", color = "black") + 
  scale_fill_continuous(name = "YPLL (2020)", low = "white", high = "red", 
                        label = scales::comma, limits = c(5000,11500)) + 
  theme(legend.position = "right") +
  labs(title = "")

state_region_map = plot_usmap(data = state_df, values = "YPLL", color = "black", include = .south_region) + #SPecify region
  scale_fill_continuous(name = "YPLL (2020)",  low = "white", high = "red",  
                        label = scales::comma, limits = c(5000,11500)) + 
  theme(legend.position = "right") +
  labs(title = "")

#Life Expectancy (low values bad --> color adjust across two plots)
# plot_usmap(data = state_df, values = "Life_Expectancy", color = "black") + 
#   scale_fill_continuous(low = "red", high = "white", name = "Life Expectancy (2020)", label = scales::comma) + 
#   theme(legend.position = "right") +
#   labs(title = "")
```

```{r}
#County
county_df = reliable_data %>%
  group_by(State) %>%
  dplyr::slice(-1) %>%
  rename(., fips = FIPS, YPLL = 'Years of Potential Life Lost Rate', Life_Expectancy = 'Life Expectancy') #%>%
  #filter(Life_Expectancy < 100, YPLL < 25000)

#YPLL (low values good)
county_map = plot_usmap(data = county_df, values = "YPLL", color = "black") + 
  scale_fill_continuous(name = "YPLL (2020)",  low = "white", high = "red",  
                        label = scales::comma) + 
  theme(legend.position = "right") +
  labs(title = "")

#Life Expectancy (low values bad --> color adjust across two plots)
# plot_usmap(data = county_df, values = "Life_Expectancy", color = "black") + 
#   scale_fill_continuous(low = "red", high = "white", name = "Life Expectancy (2020)", label = scales::comma) + 
#   theme(legend.position = "right") +
#   labs(title = "")
```

```{r}
#state level demo data
demo_response_df = reliable_data %>%
  group_by(State) %>%
  dplyr::slice(1) %>%
  ungroup(State) %>%
  rename(., state = State, all_YPLL = 'Years of Potential Life Lost Rate', Life_Expectancy = 'Life Expectancy') %>%
  select(., contains(c("YPLL Rate (", "state", "all_YPLL")))

#Check unreliable demo ypll across states
# quick = function(x){x %>% is.na() %>% sum()}
# 
# demo_response_df %>%
#   select(., contains("Unreliable")) %>%
#   apply(., MARGIN = 2, FUN = quick)

#remove unreliable demo information
index = as.numeric(is.na(demo_response_df$`YPLL Rate (AIAN) Unreliable`))
demo_response_df$`YPLL Rate (AIAN)`[which(index == 0)] = NA

#long format data
demo_response_clean_df = demo_response_df %>%
  select(., -contains("Unreliable")) %>%
  pivot_longer(cols = !state,
               names_to = "Race", values_to = "YPLL")
```

```{r}
#function to get states from region for reactive plot
expand_region = function(x){
  
  states_expanded <- if(x == "Northeast"){
    c('Connecticut', 'New Jersey', 'New York', 'Pennsylvania', 'Maine', 'Massachusetts', 'New Hampshire', 'Rhode Island', 
    'Vermont')
    
    } else if(x == "Midwest"){
      c('Illinois', 'Indiana', 'Michigan', 'Ohio', 'Wisconsin', 'Iowa', 'Kansas', 'Minnesota', 'Missouri', 'Nebraska', 
    'North Dakota', 'South Dakota')
      
      } else if(x == "South"){
  c('Delaware', 'Florida', 'Georgia', 'Maryland', 'North Carolina', 'South Carolina', 'Virginia', 'District of Columbia', 
    'West Virginia', 'Alabama', 'Kentucky', 'Mississippi', 'Tennessee', 'Arkansas', 'Louisiana', 'Oklahoma', 'Texas')
        
        } else if(x == "West"){
  c('Arizona', 'Colorado', 'Idaho', 'Montana', 'Nevada', 'New Mexico', 'Utah', 'Wyoming', 'Alaska', 'California', 'Hawaii', 
    'Oregon', 'Washington')
        } else {c()} #All case
  
  states_expanded
}
```

```{r}
#Modify: cleaned race_type inputs based on shiny list, extra: regional input flexibility
demo_map = function(df = demo_response_clean_df, Race_Type, limit_set = c(), region = '') {
  
  #Filter readable input to variable name in df
  filter_input = case_when(
    Race_Type == "All" ~ "all_YPLL",
    Race_Type == "Asian" ~ "YPLL Rate (Asian)",
    Race_Type == "Black" ~ "YPLL Rate (Black)",
    Race_Type == "Hispanic" ~ "YPLL Rate (Hispanic)",
    Race_Type == "Native American" ~ "YPLL Rate (AIAN)",
    Race_Type == "White" ~ "YPLL Rate (white)"
  )
  
  #Expand region input so it is readable in plot
  filter_region = expand_region(region)

  plot_usmap(data = filter(df, Race == filter_input), values = "YPLL", color = "black", include = filter_region) + 
  scale_fill_continuous(low = "white", high = "red", name = "YPLL", label = scales::comma, limits = limit_set) + 
  theme(legend.position = "right",
        legend.key.size = unit(10, 'mm')) +
  labs(title = "") #Race_Type

}

#Difficult to set standard limits because range varies significantly across demographic groups
demo_map(Race_Type = "Asian", region = "West")
demo_map(Race_Type = "Native American", region = "South")
demo_map(Race_Type = "Black", region = "All")
demo_map(Race_Type = "Hispanic", region = "Northeast")
demo_map(Race_Type = "White", region = "Midwest")
demo_map(Race_Type = "All")
```

## Response Percentile Ranking
```{r}
county_response_df = reliable_data %>%
  rename(., all_YPLL = 'Years of Potential Life Lost Rate') %>%
  select(., State, County, all_YPLL) %>% 
  na.omit()

ypll_hist = function(pred_value){

  value = pred_value #predicted value
  num = sum(value > county_response_df$all_YPLL) #count where value > responses
  denom = length(county_response_df$all_YPLL) #count of all responses
  percentile = round(num/denom, 2) #percentile
  per_label = paste(percentile*100, "percentile", sep = " ")

  ggplot(county_response_df, aes(x = all_YPLL)) +
    geom_histogram(col = "black", fill = "steelblue", binwidth = 500) + #stat_ecdf alternative but harder for lay to interpret
    geom_vline(xintercept = value, linetype = "dashed") +
    geom_text(aes(x = value, y = 240, label= per_label), nudge_x = 1700, colour = "steelblue", angle = 0, size = 5.5) +
    scale_x_continuous(label = scales::comma) + #the nudge lvl are weird bc im tuning it for shiny
    scale_y_continuous(breaks = seq(0,200,50)) +
    theme_bw() +
    labs(x = "YPLL", y = "Count", title = "Predicted YPLL Percentile Ranking") +
    theme(plot.title = element_text(size = 20))
}

ypll_hist(12310.13)
```


## Predictors

```{r, include=F}
#function that takes in df and varname and outputs ggplot
test_func = function(df = state_df, var_name){ 
  
  ggplot(df, aes(x = get(var_name))) + #use get to transform char to var name
    geom_histogram()
}

test_func(var_name = "Deaths")
```

```{r}
#County Level
df = reliable_data %>%
  group_by(State, County) %>%
  rename(., state = State, YPLL = 'Years of Potential Life Lost Rate') %>%
  summarise(#alcohol/drug
            drink_per = `% Excessive Drinking`, dui_per = `% Driving Deaths with Alcohol Involvement`, 
            od_num = `# Drug Overdose Deaths`, vehicle_death = `# Motor Vehicle Deaths`, obesity_rate = `% Adults with Obesity`, 
            #food/exercise
            food_index = `Food Environment Index`, inactive_per = `% Physically Inactive`, 
            exercise_access_per = `% With Access to Exercise Opportunities`, food_per = `% Food Insecure`, 
            healthy_access_per = `% Limited Access to Healthy Foods`,
            #tobacco
            smoke = `% Smokers`, 
            #sex
            std = `Chlamydia Rate`, 
            #response
            YPLL) %>% 
  na.omit()

#Remove char col

num_df = df[ , !sapply(df, is.character)]

#Linear Correlation Matrix (Response vs Predictors)
lin_cor = cor(num_df, df[,length(df)]) %>% 
  round(., 3) %>% 
  as.data.frame() %>% 
  mutate(YPLL_abs = abs(YPLL)) %>%
  arrange(desc(YPLL_abs))
```

```{r, include = F}
#Dataframe with all the predictors
#Comparing county differences if we omit all NAs
county_df = reliable_data %>%
  group_by(State, County) %>%
  rename(., state = State, YPLL = 'Years of Potential Life Lost Rate') %>%
  summarise(#alcohol/drug
            drink_per = `% Excessive Drinking`, dui_per = `% Driving Deaths with Alcohol Involvement`, 
            od_num = `# Drug Overdose Deaths`, vehicle_death = `# Motor Vehicle Deaths`, obesity_rate = `% Adults with Obesity`, 
            #food/exercise
            food_index = `Food Environment Index`, inactive_per = `% Physically Inactive`, 
            exercise_access_per = `% With Access to Exercise Opportunities`, food_per = `% Food Insecure`, 
            healthy_access_per = `% Limited Access to Healthy Foods`,
            #tobacco
            smoke = `% Smokers`, 
            #sex
            std = `Chlamydia Rate`, 
            #response
            YPLL)

setdiff(county_df, df) %>% #find rows not in subsetted df comparing to parent df
  is.na() %>% 
  colSums()
```

### Predictor Maps 

```{r}
#create state lvl df
state_pred_df = reliable_data %>%
  group_by(State) %>%
  slice(1) %>%
  rename(., state = State, YPLL = 'Years of Potential Life Lost Rate') %>%
  summarise(
            #Alcohol/Drug
            drink_per = `% Excessive Drinking`, dui_per = `% Driving Deaths with Alcohol Involvement`, 
            #food/exercise
            food_index = `Food Environment Index`, inactive_per = `% Physically Inactive`, 
            exercise_access_per = `% With Access to Exercise Opportunities`, food_per = `% Food Insecure`, 
            healthy_access_per = `% Limited Access to Healthy Foods`, obesity_rate = `% Adults with Obesity`,
            #tobacco
            smoke = `% Smokers`,
            #sex
            std = `Chlamydia Rate`,
            #response
            YPLL,
            #Cluster
            county_class = as.factor(county_class),
            #Population
            Population) %>%
  na.omit()

#Create parsing filters for pred map function
#simple --> label --> varname
pred_options = c("Excessive Drinking", "DUI", "Food Index", "Inactivity", "Exercise Opportunity", "Food Insecurity", "Healthy Food Access",
                 "Obesity", "Smoking", "STI") %>% sort()

simple_to_label = function(x){
  
  label = case_when(
    x == "Excessive Drinking" ~ "% Excessive Drinking",
    x == "DUI" ~ "% Driving Deaths with Alcohol Involvement",
    x == "Food Index" ~ "Food Environment Index",
    x == "Inactivity" ~ "% Physically Inactive",
    x == "Exercise Opportunity" ~ "% With Access to Exercise Opportunities",
    x == "Food Insecurity" ~ "% Food Insecure",
    x == "Healthy Food Access" ~ "% Limited Access to Healthy Foods",
    x == "Obesity" ~ "% Adults with Obesity",
    x == "Smoking" ~ "% Smokers",
    x == "STI"~ "Chlamydia Rate"
  )
  
}


label_to_varname = function(x){
  
    varname = case_when(
    x == "% Excessive Drinking" ~ "drink_per",
    x == "% Driving Deaths with Alcohol Involvement" ~ "dui_per",
    x == "Food Environment Index" ~ "food_index",
    x == "% Physically Inactive" ~ "inactive_per",
    x == "% With Access to Exercise Opportunities" ~ "exercise_access_per",
    x == "% Food Insecure" ~ "food_per",
    x == "% Limited Access to Healthy Foods" ~ "healthy_access_per",
    x == "% Adults with Obesity" ~ "obesity_rate",
    x == "% Smokers" ~ "smoke",
    x == "Chlamydia Rate" ~ "std"
  )
    varname
}
```


```{r}
pred_map = function(df = state_pred_df, pred, limit_set = c(), region = '') {
  
  #Simple to label
  label_x = simple_to_label(pred)
  
  #Label to varname
  varname = label_to_varname(label_x)
  
  #Expand region input so it is readable in plot
  filter_region = expand_region(region)

  plot_usmap(data = df, values = varname, color = "black", include = filter_region) + 
  scale_fill_continuous(low = "white", high = "red", name = label_x, label = scales::comma, limits = limit_set) + 
  theme(legend.position = "right",
        legend.key.size = unit(10, 'mm')) +
  labs(title = "") #Race_Type

}

pred_map(pred = "Excessive Drinking", region = "West")
pred_map(pred = "DUI", region = "South")
pred_map(pred = "Food Index", region = "Northeast")
pred_map(pred = "Inactivity", region = "Midwest")
pred_map(pred = "STI", region = "All")
```

```{r}
county_pred_df = reliable_data %>%
  group_by(State, County) %>%
  rename(., state = State, YPLL = 'Years of Potential Life Lost Rate') %>%
  summarise(
            #Alcohol/Drug
            drink_per = `% Excessive Drinking`, dui_per = `% Driving Deaths with Alcohol Involvement`, 
            #od_num = `# Drug Overdose Deaths`, vehicle_death = `# Motor Vehicle Deaths`, #remove
            #food/exercise
            food_index = `Food Environment Index`, inactive_per = `% Physically Inactive`, 
            exercise_access_per = `% With Access to Exercise Opportunities`, food_per = `% Food Insecure`, 
            healthy_access_per = `% Limited Access to Healthy Foods`, obesity_rate = `% Adults with Obesity`,
            #tobacco
            smoke = `% Smokers`,
            #sex
            std = `Chlamydia Rate`,
            #response
            YPLL,
            #Cluster
            county_class = as.factor(county_class),
            #Population
            Population) %>%
  na.omit() 

county_num_df = county_pred_df[ ,!sapply(county_pred_df, is.character)] %>%
  relocate(YPLL, .before = drink_per)

# end = length(county_num_df)
# half = (end/2) %>% ceiling()
# library(GGally)
# ggpairs(county_num_df[,c(1:half)])
# ggpairs(county_num_df[,c(1,half:end)])
# cor(county_num_df[,-c(end-1,)]) %>% round(2)
```

### Bivariate Analysis
```{r}
#response vs pred
ypll_smoke = ggplot(county_pred_df, aes(y = YPLL, x = smoke)) +
  geom_point(alpha = 0.25) +
  theme_bw() +
  geom_smooth(se = F) +
  labs(y = "YPLL", x = "% Smokers") #title = "YPLL Increases As Smoking Prevelance Increases"

ypll_drink = ggplot(county_pred_df, aes(y = YPLL, x = drink_per)) +
  geom_point(alpha = 0.25) +
  theme_bw() +
  geom_smooth(se = F) +
  labs(y = "YPLL", x = "% Excessive Drinking") #title = "YPLL Decreases As Smoking Prevelance Increases"

ypll_food = ggplot(county_pred_df, aes(y = YPLL, x = food_per)) +
  geom_point(alpha = 0.25) +
  theme_bw() +
  geom_smooth(se = F) +
  labs(y = "YPLL", x = "% Food Insecure") #title = "YPLL Increases As Food Insecurity Increases"

ypll_obesity = ggplot(county_pred_df, aes(y = YPLL, x = obesity_rate)) +
  geom_point(alpha = 0.25) +
  theme_bw() +
  geom_smooth(se = F) +
  labs(y = "YPLL", x = "% Adults with Obesity") #title = "YPLL Increases As Obesity Levels Increases"

ypll_countyclass = ggplot(county_pred_df, aes(y = YPLL, x = county_class)) +
  geom_violin() +
  geom_boxplot(width = .2, alpha = 0.25) +
  scale_x_discrete(limits = c("Urban", "Suburban", "Rural")) +
  theme_bw() +
  labs(y = "YPLL", x = "County Type") #title = "YPLL Variability Increase as County Population Decreases"

library(cowplot)
plot_grid(ypll_drink, ypll_smoke,ypll_food, ypll_obesity, 
          ncol = 2, labels = c('A', 'B', 'C', 'D'), label_fontface = 'bold',label_size = 11)
```

```{r}
#YPLL vs Predictors colored by county class
ypll_countyclass #violin overlay on box plot, variability limitation in YPLL calculation for low pop counties
```

```{r}
#Pred vs pred
food_obesity = ggplot(county_pred_df, aes(y = food_per, x = obesity_rate)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = F) +
  theme_bw() +
  labs(y = "% Food Insecure", x = "% Adults with Obesity") #title = "Positive Relationship Between Obesity and Food Insecurity"

food_drink = ggplot(county_pred_df, aes(y = food_per, x = drink_per)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = F) +
  theme_bw() +
  labs(y = "% Food Insecure", x = "% Excessive Drinking") #title = "Inverse Relationship Between Food Insecurity and Excessive Drinking"

smoke_dui = ggplot(county_pred_df, aes(y = smoke, x = dui_per)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = F) +
  theme_bw() +
  labs(y = "% Smokers", x = "% Driving Deaths with Alcohol Involvement") #title = "No Relationship Between Smoking and Alcohol Driving Deaths"

plot_grid(food_obesity, food_drink, smoke_dui, 
          ncol = 2, labels = c('A', 'B', 'C'), label_fontface = 'bold', label_size = 11)
```

# Modeling

## Variable Selection

Lasso, Ridge, Elastic Net (Deal with correlation especially across diet/exercise related variables)

```{r}
library(glmnet)
library(broom)

#Regularization Modeling
#stratified cv (e.g., randomly assign 1:5 across all counties within each state)
strat_cv = function(df_input, k = 5, seed = 123){
  
  set.seed(seed)
  state_list = unique(df_input$state)
  strat_df = data.frame()

  for (i in 1:length(state_list)){
    df = filter(df_input, state == state_list[i])
    new_df = mutate(df, test_fold = sample(1:k, replace = T, size = nrow(df)))
    strat_df = rbind(strat_df, new_df)
  }
  
  strat_df %>% ungroup() %>% select(-c(state, County))
}

cv_df <- strat_cv(df_input = county_pred_df) %>%
    mutate(urban = ifelse(county_class == 'Urban', 1, 0), #transform county_class to dummy
            suburban = ifelse(county_class == 'Suburban', 1, 0),
            rural = ifelse(county_class == 'Rural', 1, 0)) %>%
  select(-c(county_class, Population))

holdout_predictions <- 
  map_dfr(unique(cv_df$test_fold), 
          
          function(holdout) {
            # Separate test and training data:
            test_data <- cv_df %>% filter(test_fold == holdout)
            train_data <- cv_df %>% filter(test_fold != holdout)
            
            # Repeat for matrices
            test_x <- as.matrix(dplyr::select(test_data, -YPLL))
            train_x <- as.matrix(dplyr::select(train_data, -YPLL))

            # Train models:
            lm_model <- lm(YPLL ~ ., data = train_data)
            ridge_model <- cv.glmnet(train_x, train_data$YPLL, alpha = 0)
            lasso_model <- cv.glmnet(train_x, train_data$YPLL, alpha = 1)
            en_model <- cv.glmnet(train_x, train_data$YPLL, alpha = .5)

            # Return tibble of holdout results:
            tibble(lm_preds = predict(lm_model, newdata = test_data), #linear model
                   ridge_preds = as.numeric(predict(ridge_model, newx = test_x)), #ridge
                   lasso_preds = as.numeric(predict(lasso_model, newx = test_x)), #lasso
                   en_preds = as.numeric(predict(en_model, newx = test_x)), #elastic net
                   test_actual = test_data$YPLL, test_fold = holdout) 
          })

reg_cv = holdout_predictions %>%
  pivot_longer(lm_preds:en_preds, 
               names_to = "type", values_to = "test_preds") %>%
  group_by(type, test_fold) %>%
  summarize(rmse = sqrt(mean((test_actual - test_preds)^2))) %>% 
  mutate(type = recode(type, 'en_preds' = 'Elastic Net' , 'lasso_preds' =  'Lasso','ridge_preds' = 'Ridge', 'lm_preds' = 'Linear')) %>%
  ggplot(aes(x = type, y = rmse)) + 
  geom_point() + 
  scale_x_discrete(limits = c("Linear", "Lasso","Elastic Net", "Ridge")) + 
  theme_bw() +
  stat_summary(fun = mean, geom = "point", 
               color = "red") + 
  stat_summary(fun.data = mean_se, geom = "errorbar",
               color = "red") +
  labs(y = "RMSE", x = "Model Type", title = "Linear Regression Outperforms Regularization Techniques")
```

## Model Design

### Linear Regression Model

```{r}
library(ggfortify)

#Aggregated data: Underestimates variability, Inflated SE, Inflated R^2
lm_mod = lm(YPLL ~ std + smoke + food_per + drink_per + county_class, data = county_num_df)

#autoplot(lm_mod, ncol = 2) + theme_bw() #mod assumptions

#summary(lm_mod) #Output
```

### GAM

```{r}
library(caret)
library(mgcv)
```
#XGBoost
```{r}
library(xgboost)
library(vip)
xgboost_tune_grid <- expand.grid(nrounds = seq(from = 20, to = 200, by = 20),
                                 eta = c(0.025, 0.05, 0.1, 0.3), gamma = 0, #seq(0, 4, 2)
                                 max_depth = 3, colsample_bytree = 0.9, #c(1, 2, 3, 4), seq(0.8, 1, 0.1)
                                 min_child_weight = 1, subsample = 0.9) #seq(0.8, 1, 0.1)

xgboost_tune_control <- trainControl(method = "cv", number = 5, verboseIter = FALSE)

set.seed(1937)

xg_df = strat_cv(df_input = county_pred_df) %>%
    mutate(urban = ifelse(county_class == 'Urban', 1, 0), #transform county_class to dummy
            suburban = ifelse(county_class == 'Suburban', 1, 0),
            rural = ifelse(county_class == 'Rural', 1, 0)) %>%
  select(-c(county_class, Population, test_fold))

xgb_tune <- train(x = as.matrix(dplyr::select(xg_df, -YPLL)),
                  y = xg_df$YPLL, trControl = xgboost_tune_control,
                  tuneGrid = xgboost_tune_grid, 
                  objective = "reg:squarederror", method = "xgbTree",
                  verbose = F)

xgb_fit_final <- xgboost(data = as.matrix(dplyr::select(xg_df, -YPLL)),
                         label = xg_df$YPLL, objective = "reg:squarederror",
                         nrounds = xgb_tune$bestTune$nrounds,
                         params = as.list(dplyr::select(xgb_tune$bestTune, -nrounds)), 
                         verbose = 0)

vip(xgb_fit_final) + theme_bw()

xgb_tune$finalModel
```

## Model Validation

```{r}
#GAM Modeling
#stratified k fold
strat_cv_df = strat_cv(df_input = county_pred_df)

#Calculate holdout predictions
holdout_predictions <- 
  map_dfr(unique(strat_cv_df$test_fold), 
          
          function(holdout) {
            # Separate test and training data:
            test_data <- strat_cv_df %>% filter(test_fold == holdout)
            train_data <- strat_cv_df %>% filter(test_fold != holdout)
            
            xg_train <- train_data %>% 
              mutate(urban = ifelse(county_class == 'Urban', 1, 0), #transform county_class to dummy
                      suburban = ifelse(county_class == 'Suburban', 1, 0),
                      rural = ifelse(county_class == 'Rural', 1, 0)) %>%
              select(-c(county_class, Population, test_fold))
            
            xg_test <- test_data %>% 
              mutate(urban = ifelse(county_class == 'Urban', 1, 0), #transform county_class to dummy
                      suburban = ifelse(county_class == 'Suburban', 1, 0),
                      rural = ifelse(county_class == 'Rural', 1, 0)) %>%
              select(-c(county_class, Population, test_fold))
          
            # Train models:
            lm_model <- lm(YPLL ~ ., data = train_data)
            
            xgb_model <- xgboost(data = as.matrix(dplyr::select(xg_train, -YPLL)),
                         label = train_data$YPLL, objective = "reg:squarederror",
                         nrounds = xgb_tune$bestTune$nrounds,
                         params = as.list(dplyr::select(xgb_tune$bestTune, -nrounds)), 
                         verbose = 0)
            
            gam_model <- gam(YPLL ~ s(drink_per) + s(dui_per) + s(food_index) + s(inactive_per) + s(exercise_access_per) + 
                             s(food_per) + s(healthy_access_per) + s(obesity_rate) + s(smoke) + s(std) + as.factor(county_class), 
                            data = train_data, 
                            method = "REML")

            simple_gam_model <-  gam(YPLL ~ s(inactive_per) + s(obesity_rate) + s(food_per) + s(drink_per) + s(smoke) + s(std) +
                                     as.factor(county_class),
                                     data = train_data, 
                                     method = "REML")

            # Return tibble of holdout results:
            tibble(lm_preds = predict(lm_model, newdata = test_data), #linear model
                   xgb_preds = predict(xgb_model, newdata = as.matrix(dplyr::select(xg_test, -YPLL))),
                   gam_preds = predict(gam_model, newdata = test_data), #gam model
                   simple_gam_preds = predict(simple_gam_model, newdata = test_data),
                   test_actual = test_data$YPLL, test_fold = holdout)
          })

#error notes: RMSE and MAE put the error on response scale, such that our model on average is MAE/RMSE units away from the actual response
gam_cv = holdout_predictions %>%
  pivot_longer(lm_preds:simple_gam_preds, 
               names_to = "type", values_to = "test_preds") %>%
  group_by(type, test_fold) %>%
  summarize(rmse = sqrt(mean((test_actual - test_preds)^2)),
            mae = mean(abs(test_actual - test_preds))) %>%
  mutate(type = recode(type, 'gam_preds' = 'Full GAM' , 'simple_gam_preds' =  'Simple GAM', 'xgb_preds' = 'XGBoost',
                       'lm_preds' = 'Linear')) %>%
  ggplot(aes(x = type, y = rmse)) + 
  geom_point() + 
  scale_x_discrete(limits = c("Simple GAM", "XGBoost", "Full GAM", "Linear")) + 
  theme_bw() +
  stat_summary(fun = mean, geom = "point", 
               color = "red") + 
  stat_summary(fun.data = mean_se, geom = "errorbar",
               color = "red") +
  labs(y = "RMSE", x = "Model Type", title = "Simple GAM Outperforms Full GAM and Linear Models")
```

Simple gam (7 features) performs outperforms gam with all 11 features within intervals. For simplicity and performance in potential Shiny App interface simple gam seems to be the way to go.

Interesting is that population instead of county_class outperforms very slightly, which is to be expected especially in GAM bc we lose some variability in information when transforming population to 3 county classes

```{r}
#Fitted vs Residuals for final model
#Final model df
final_df = reliable_data %>%
  group_by(State, County) %>%
  rename(., state = State, YPLL = 'Years of Potential Life Lost Rate') %>%
  summarise(
            #Alcohol/Drug
            drink_per = `% Excessive Drinking`, 
            #food/exercise
            inactive_per = `% Physically Inactive`, 
            food_per = `% Food Insecure`, 
            obesity_rate = `% Adults with Obesity`,
            #tobacco
            smoke = `% Smokers`,
            #sex
            std = `Chlamydia Rate`,
            #response
            YPLL,
            #Cluster
            county_class = as.factor(county_class),
            #Population
            Population) %>%
  na.omit()

#Train/Test set
set.seed(1)
parts = createDataPartition(final_df$YPLL, p = .7, list = F) #70/30 split
train = final_df[parts, ]
test = final_df[-parts, ]

plot_gam_model <- gam(YPLL ~ s(inactive_per) + s(obesity_rate) + s(food_per) + s(drink_per) + s(smoke) + s(std) +
                         as.factor(county_class), #alternative: s(Population)
                      data = train, 
                      method = "REML")

y_actual = test$YPLL
y_fitted = predict(plot_gam_model, test)

fin_mod = data.frame(y_actual, y_fitted) %>%
  ggplot(., aes(x = y_actual, y = y_fitted)) +
  geom_point(alpha = 0.25) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  coord_equal() +
  scale_x_continuous(breaks = seq(0, 25000, 5000)) +
  theme_bw() +
  geom_smooth(se = F) +
  labs(x = "Actual", y = "Predicted")
```

```{r}
#Finalized shiny model
#Note: The F statistic explains predictors impact on output response
final_gam_model <- gam(YPLL ~ s(inactive_per) + s(obesity_rate) + s(food_per) + s(drink_per) + s(smoke) + s(std) +
                         as.factor(county_class), #alternative: s(Population)
                      data = final_df, 
                      method = "REML")

#summary(final_gam_model)
#plot(final_gam_model)
```

```{r}
#Allegheny Info
filter(final_df, County == "Allegheny")

# Pennsylvania,	Allegheny	
# county_class: Urban
# inactive_per: 24
# obesity_rate: 29
# food_per: 11
# drink_per: 23
# smoke: 19
# std: 520

# YPLL: 7699.26	
```

# To Do: (1) VIP/Inference/Assumptions for GAM, (2) Publish Shiny 
# Shiny App

```{r}
library(shiny)
library(shinyBS)
library(bslib) #theme
library(hablar) #retype()
library(formattable)#comma()

#CSS: https://unleash-shiny.rinterface.com/beautify-css.html

county_options = c("Urban", "Suburban", "Rural") %>% sort()
demo_options = c("All", "Asian", "Black", "Hispanic", "Native American", "White") %>% sort()
region_options = c('All',"Northeast", "Midwest", "South", "West") %>% sort()
df_names = c("county_class", "inactive_per", "obesity_rate", "food_per", "drink_per", "smoke", "std")
pred_options = c("Excessive Drinking", "Inactivity", "Food Insecurity", "Obesity", "Smoking", "STI") %>% sort()
```

```{r}
#Modify tabpanel title height x width (CSS needed)
ui = navbarPage(title = div(img(
  
  src = "https://www.logolynx.com/images/logolynx/e7/e7ec1b0c8c294ff754d8e20c5636e6ae.png", height = 60, width = 210), ""),
  
  windowTitle = "Optum Bridges",
  
  #theme = bs_theme(bootswatch = "lux"), #theme turns off bstooltips
  
  inverse = T,
  
          tags$head(
           tags$style(HTML('.navbar-nav > li > a, .navbar-brand {
                            padding-top:18px !important; 
                            padding-bottom:0px !important;
                            height: 90px;
                            }')) #Alter height of navbarpage using CSS
         ),
                
      tabPanel(#classifying css class with text attributes that can be called
        tags$style(".first-p {color: white;
                   padding-top:18px !important;
                   font-size: 16px;
                   font-weight: 500;}"
                   ), 
        
        title = p(class = "first-p", "Predictive Modeling"),
    
           fluidPage(
            
             titlePanel(title = h2("Premature Deaths Prediction", align = "center")),
              
             sidebarLayout(
               
             sidebarPanel( 
                selectInput(inputId = "county_class", label = "County Type:", choices = county_options),
                numericInput(inputId = "inactive_per", label = "Inactivity %:", value = 0, min = 0, max = 100, step = 0.1),
                numericInput(inputId = "obesity_rate", label = "Obesity %:", value = 0, min = 0, max = 100, step = 0.1),
                numericInput(inputId = "food_per", label = "Food Insecure %:", value = 0, min = 0, max = 100, step = 0.1),
                numericInput(inputId = "drink_per", label = "Excessive Drinking %:", value = 0, min = 0, max = 100, step = 0.1),
                numericInput(inputId = "smoke", label = "Smoking %:", value = 0, min = 0, max = 100, step = 0.1),
                numericInput(inputId = "std", label = "Chlamydia Cases:", value = 0, min = 0, max = 5000, step = 0.1),
                
                actionButton(inputId = "go", label = "Predict"),
                
                bsTooltip(id = "county_class", title = "County Description"),
                bsTooltip(id = "inactive_per", title = "% of Adults Reporting No Leisure Time Physical Activity"),
                bsTooltip(id = "obesity_rate", title = "% of Adults with BMI > 30"),
                bsTooltip(id = "food_per", title = "% of Population With Poor Access to Food"),
                bsTooltip(id = "drink_per", title = "% of Adults Reporting Binge/Heavy Drinking"),
                bsTooltip(id = "smoke", title = "% of Adults Smoking Tobacco"),
                bsTooltip(id = "std", title = "Number of Chlamydia Cases per 100k")
             ),
             
            mainPanel(br(), #manually add line break to create space for plot
                      plotOutput("plot1"),
                      uiOutput("text1"),
              tags$head(tags$style("#text1{color: black;
                                 font-size: 22px;
                                 line-height: 17.0;
                                 text-indent: 37px;
                                 }"
                         )))

             ))),
  
    navbarMenu(title = p(class = "first-p", "Interactive Map"),
               
    tabPanel(title = "Premature Deaths",
             
    fluidPage(
        
        titlePanel(title = h2("Premature Deaths Map", align = "center")),
        
        sidebarLayout(
        sidebarPanel( 
            selectInput(inputId = "demo", label = "Race:", choices = demo_options),
            selectInput(inputId = "region", label = "Region:", choices = region_options), #optional addition
            
            actionButton(inputId = "go_response", label = "Update")
            ),
      
        mainPanel(
          plotOutput("plot2"))
        ))         
             
  ),

    tabPanel(title = "Health Behaviors",
             
    fluidPage(
        
        titlePanel(title = h2("Health Behaviors Map", align = "center")),
        
        sidebarLayout(
        sidebarPanel( 
            selectInput(inputId = "pred", label = "Health Behavior:", choices = pred_options),
            selectInput(inputId = "region_pred", label = "Region:", choices = region_options), #optional addition
            
            actionButton(inputId = "go_pred", label = "Update")
            ),
      
        mainPanel(
          plotOutput("plot3"))
        ))         
             
  )
))
```

```{r}
server = function(input,output){
  
  #store inputs
  x_value <- reactive({
    c(input$county_class, input$inactive_per, input$obesity_rate, input$food_per, input$drink_per, input$smoke, input$std)
  })
  
  #convert inputs to dataframe
  shiny_df <- reactive({
    new_table = data.frame((matrix(nrow = 1, data = x_value()))) %>% retype()
    colnames(new_table) = df_names #get appropriate colnames
    new_table
  })
  
  #Get predictions
  shiny_pred <- reactive({
    predict(final_gam_model, newdata = shiny_df()) %>% comma(., digits = 2) #add comma to thousand and round by 2
  })
  
  #Text output
  pred_reactive <- eventReactive(input$go, #put the expected YPLL text together
    {
      paste("The expected county YPLL per 100k is:", shiny_pred(), "Years")
  })
  
  output$text1 <- renderUI({pred_reactive()})
  
  #Plot output
  ypll_plot_1 <- eventReactive(input$go, {
    ypll_hist(pred_value = as.numeric(shiny_pred()))
    })
    
  output$plot1 <- renderPlot({
    ypll_plot_1()
  }, height = 550, width = 900)
  
  response_plot <- eventReactive(input$go_response, {
    demo_map(Race_Type = input$demo, region = input$region)
    })
  
  output$plot2 <- renderPlot({
    response_plot()
  }, height = 500, width = 800)
  
  pred_plot <- eventReactive(input$go_pred, {
    pred_map(pred = input$pred, region = input$region_pred)
    })
  
  output$plot3 <- renderPlot({
    pred_plot()
  }, height = 500, width = 800)
}
```

```{r}
#run app
shinyApp(ui = ui, server = server)
```
