---
title: "Premature Death Prediction Model Using Health Behavior Data"
author: 'Donald Dinerman, Masara Greenidge, Suron Lockett'
date: '2022-07-22'
output: html_document
---

```{r setup, include=FALSE}
## Question: use geom_smooth? if so lm method or gam?, what to add for results section? GAM inference?
#Going to run all code above HTML, save output as objects and then call objects in html code chunks
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

```{r}
#load libraries
library(tidyverse)
library(readxl)
library(usmap)
library(ggplot2)
library(glmnet)
library(broom)
library(caret)
library(mgcv)
library(cowplot)
```

```{r}
#load data
data = read_xlsx("Data/CountyHealthRankingsData2022.xlsx", 4, skip = 1)

add_data = read_xlsx("Data/CountyHealthRankingsData2022.xlsx", 6, skip = 1)
```

```{r}
#process data
new_data = select(data, -contains(c("CI", "Quartile"), ignore.case = F))
new_add_data = select(add_data, -contains(c("CI", "Quartile"), ignore.case = F))
comb_data = inner_join(new_data, new_add_data, by = c("FIPS", "State", "County"))

reliable_data = comb_data %>%
  mutate(na_id = as.numeric(is.na(comb_data$Unreliable...4))) %>% #binary to is.na
  filter(na_id == 1) #keep only rows w NA (reliable rows)

sub_upper_lim = 200000
sub_lower_lim = 50000

reliable_data = reliable_data %>%
  mutate(county_class = 
           case_when(Population < (sub_upper_lim) & Population > (sub_lower_lim) ~ "Suburban",
                     Population >= (sub_upper_lim) ~ "Urban", 
                     TRUE ~ "Rural")) #%>% select(State, County, county_class, Population)
```

```{r}
#State
state_df = reliable_data %>%
  group_by(State) %>%
  dplyr::slice(1) %>%
  rename(., state = State, YPLL = 'Years of Potential Life Lost Rate', Life_Expectancy = 'Life Expectancy')

#YPLL (low values good)
ypll_map = plot_usmap(data = state_df, values = "YPLL", color = "black") + 
  scale_fill_continuous(name = "YPLL", low = "white", high = "red", 
                        label = scales::comma, limits = c(5000,11500)) + 
  theme(legend.position = "right") +
  labs(title = "")
```

```{r}
county_pred_df = reliable_data %>%
  group_by(State, County) %>%
  rename(., state = State, YPLL = 'Years of Potential Life Lost Rate') %>%
  summarise(
            #Alcohol/Drug
            drink_per = `% Excessive Drinking`, dui_per = `% Driving Deaths with Alcohol Involvement`, 
            #food/exercise
            food_index = `Food Environment Index`, inactive_per = `% Physically Inactive`, 
            exercise_access_per = `% With Access to Exercise Opportunities`, food_per = `% Food Insecure`, 
            healthy_access_per = `% Limited Access to Healthy Foods`, obesity_rate = `% Adults with Obesity`,
            #tobacco
            smoke = `% Smokers`,
            #sex
            std = `Chlamydia Rate`,
            #response
            YPLL,
            #Cluster
            county_class = as.factor(county_class),
            #Population
            Population) %>%
  na.omit()

ypll_smoke = ggplot(county_pred_df, aes(y = YPLL, x = smoke)) +
  geom_point(alpha = 0.25) +
  scale_y_continuous(label = scales::comma) +
  theme_bw() +
  geom_smooth(se = F) +
  labs(y = "YPLL", x = "% Smokers") #title = "YPLL Increases As Smoking Prevelance Increases"

ypll_drink = ggplot(county_pred_df, aes(y = YPLL, x = drink_per)) +
  geom_point(alpha = 0.25) +
  scale_y_continuous(label = scales::comma) +
  theme_bw() +
  geom_smooth(se = F) +
  labs(y = "YPLL", x = "% Excessive Drinking") #title = "YPLL Decreases As Smoking Prevelance Increases"

ypll_food = ggplot(county_pred_df, aes(y = YPLL, x = food_per)) +
  geom_point(alpha = 0.25) +
  scale_y_continuous(label = scales::comma) +
  theme_bw() +
  geom_smooth(se = F) +
  labs(y = "YPLL", x = "% Food Insecure") #title = "YPLL Increases As Food Insecurity Increases"

ypll_obesity = ggplot(county_pred_df, aes(y = YPLL, x = obesity_rate)) +
  geom_point(alpha = 0.25) +
  scale_y_continuous(label = scales::comma) +
  theme_bw() +
  geom_smooth(se = F) +
  labs(y = "YPLL", x = "% Adults with Obesity") #title = "YPLL Increases As Obesity Levels Increases"

ypll_countyclass = ggplot(county_pred_df, aes(y = YPLL, x = county_class)) +
  geom_violin() +
  geom_boxplot(width = .2, alpha = 0.25) +
  scale_x_discrete(limits = c("Urban", "Suburban", "Rural")) +
  scale_y_continuous(label = scales::comma) +
  theme_bw() +
  labs(y = "YPLL", x = "County Type") #title = "YPLL Variability Increase as County Population Decreases"
```

```{r}
food_obesity = ggplot(county_pred_df, aes(y = food_per, x = obesity_rate)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = F) +
  theme_bw() +
  labs(y = "% Food Insecure", x = "% Adults with Obesity") #title = "Positive Relationship Between Obesity and Food Insecurity"

food_drink = ggplot(county_pred_df, aes(y = food_per, x = drink_per)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = F) +
  theme_bw() +
  labs(y = "% Food Insecure", x = "% Excessive Drinking") #title = "Inverse Relationship Between Food Insecurity and Excessive Drinking"

smoke_dui = ggplot(county_pred_df, aes(y = smoke, x = dui_per)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = F) +
  theme_bw() +
  labs(y = "% Smokers", x = "% Driving Deaths with Alcohol Involvement") #title = "No Relationship Between Smoking and Alcohol Driving Deaths"
```

```{r}
#Regularization Modeling
#stratified cv
strat_cv = function(df_input, k = 5, seed = 123){
  
  set.seed(seed)
  state_list = unique(df_input$state)
  strat_df = data.frame()

  for (i in 1:length(state_list)){
    df = filter(df_input, state == state_list[i])
    new_df = mutate(df, test_fold = sample(1:k, replace = T, size = nrow(df)))
    strat_df = rbind(strat_df, new_df)
  }
  
  strat_df %>% ungroup() %>% select(-c(state, County))
}

cv_df <- strat_cv(df_input = county_pred_df) %>%
    mutate(urban = ifelse(county_class == 'Urban', 1, 0), #transform county_class to dummy
            suburban = ifelse(county_class == 'Suburban', 1, 0),
            rural = ifelse(county_class == 'Rural', 1, 0)) %>%
  select(-c(county_class, Population))

holdout_predictions <- 
  map_dfr(unique(cv_df$test_fold), 
          
          function(holdout) {
            # Separate test and training data:
            test_data <- cv_df %>% filter(test_fold == holdout)
            train_data <- cv_df %>% filter(test_fold != holdout)
            
            # Repeat for matrices
            test_x <- as.matrix(dplyr::select(test_data, -YPLL))
            train_x <- as.matrix(dplyr::select(train_data, -YPLL))

            # Train models:
            lm_model <- lm(YPLL ~ ., data = train_data)
            ridge_model <- cv.glmnet(train_x, train_data$YPLL, alpha = 0)
            lasso_model <- cv.glmnet(train_x, train_data$YPLL, alpha = 1)
            en_model <- cv.glmnet(train_x, train_data$YPLL, alpha = .5)

            # Return tibble of holdout results:
            tibble(lm_preds = predict(lm_model, newdata = test_data), #linear model
                   ridge_preds = as.numeric(predict(ridge_model, newx = test_x)), #ridge
                   lasso_preds = as.numeric(predict(lasso_model, newx = test_x)), #lasso
                   en_preds = as.numeric(predict(en_model, newx = test_x)), #elastic net
                   test_actual = test_data$YPLL, test_fold = holdout) 
          })

reg_cv = holdout_predictions %>%
  pivot_longer(lm_preds:en_preds, 
               names_to = "type", values_to = "test_preds") %>%
  group_by(type, test_fold) %>%
  summarize(rmse = sqrt(mean((test_actual - test_preds)^2))) %>% 
  mutate(type = recode(type, 'en_preds' = 'Elastic Net' , 'lasso_preds' =  'Lasso','ridge_preds' = 'Ridge', 'lm_preds' = 'Linear')) %>%
  ggplot(aes(x = type, y = rmse)) + 
  geom_point() + 
  scale_x_discrete(limits = c("Linear", "Lasso","Elastic Net", "Ridge")) + 
  scale_y_continuous(label = scales::comma) +
  theme_bw() +
  stat_summary(fun = mean, geom = "point", 
               color = "red") + 
  stat_summary(fun.data = mean_se, geom = "errorbar",
               color = "red") +
  labs(y = "RMSE", x = "Model Type", title = "Linear Regression Outperforms Regularization Techniques")
```

```{r}
#GAM Modeling
#stratified k fold
strat_cv_df = strat_cv(df_input = county_pred_df)

#Calculate holdout predictions
holdout_predictions <- 
  map_dfr(unique(strat_cv_df$test_fold), 
          
          function(holdout) {
            # Separate test and training data:
            test_data <- strat_cv_df %>% filter(test_fold == holdout)
            train_data <- strat_cv_df %>% filter(test_fold != holdout)
          
            # Train models:
            lm_model <- lm(YPLL ~ ., data = train_data)
            
            gam_model <- gam(YPLL ~ s(drink_per) + s(dui_per) + s(food_index) + s(inactive_per) + s(exercise_access_per) + s(food_per) +
                            s(healthy_access_per) + s(obesity_rate) + s(smoke) + s(std) + as.factor(county_class), 
                            data = train_data, 
                            method = "REML")

            simple_gam_model <-  gam(YPLL ~ s(inactive_per) + s(obesity_rate) + s(food_per) + s(drink_per) + s(smoke) + s(std) +
                                     as.factor(county_class),
                                     data = train_data, 
                                     method = "REML")

            # Return tibble of holdout results:
            tibble(lm_preds = predict(lm_model, newdata = test_data), #linear model
                   gam_preds = predict(gam_model, newdata = test_data), #gam model
                   simple_gam_preds = predict(simple_gam_model, newdata = test_data),
                   test_actual = test_data$YPLL, test_fold = holdout)
          })

gam_cv = holdout_predictions %>%
  pivot_longer(lm_preds:simple_gam_preds, 
               names_to = "type", values_to = "test_preds") %>%
  group_by(type, test_fold) %>%
  summarize(rmse = sqrt(mean((test_actual - test_preds)^2))) %>%
  mutate(type = recode(type, 'gam_preds' = 'Full GAM' , 'simple_gam_preds' =  'Simple GAM', 'lm_preds' = 'Linear')) %>%
  ggplot(aes(x = type, y = rmse)) + 
  geom_point() + 
  scale_x_discrete(limits = c("Simple GAM", "Full GAM", "Linear")) + 
  scale_y_continuous(label = scales::comma) +
  theme_bw() +
  stat_summary(fun = mean, geom = "point", 
               color = "red") + 
  stat_summary(fun.data = mean_se, geom = "errorbar",
               color = "red") +
  labs(y = "RMSE", x = "Model Type", title = "Simple GAM Outperforms Full GAM and Linear Models")
```

```{r}
final_df = reliable_data %>%
  group_by(State, County) %>%
  rename(., state = State, YPLL = 'Years of Potential Life Lost Rate') %>%
  summarise(
            #Alcohol/Drug
            drink_per = `% Excessive Drinking`, 
            #food/exercise
            inactive_per = `% Physically Inactive`, 
            food_per = `% Food Insecure`, 
            obesity_rate = `% Adults with Obesity`,
            #tobacco
            smoke = `% Smokers`,
            #sex
            std = `Chlamydia Rate`,
            #response
            YPLL,
            #Cluster
            county_class = as.factor(county_class),
            #Population
            Population) %>%
  na.omit()

#Train/Test set
set.seed(1)
parts = createDataPartition(final_df$YPLL, p = .7, list = F) #70/30 split
train = final_df[parts, ]
test = final_df[-parts, ]

plot_gam_model <- gam(YPLL ~ s(inactive_per) + s(obesity_rate) + s(food_per) + s(drink_per) + s(smoke) + s(std) +
                         as.factor(county_class), #alternative: s(Population)
                      data = train, 
                      method = "REML")

y_actual = test$YPLL
y_fitted = predict(plot_gam_model, test)

fin_mod = data.frame(y_actual, y_fitted) %>%
  ggplot(., aes(x = y_actual, y = y_fitted)) +
  geom_point(alpha = 0.25) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  coord_equal() +
  scale_x_continuous(breaks = seq(0, 25000, 5000), label = scales::comma) +
  scale_y_continuous(label = scales::comma) +
  theme_bw() +
  geom_smooth(se = F) +
  labs(x = "Actual", y = "Predicted")
```

```{r, include=F}
#how to do VIP for gam
```

## Introduction

- Question: How can we predict premature deaths using county level health behavior data?

- Question: Which particular health behaviors are most influential in modeling premature deaths?

The data given from County Health Rankings allowed us to analyze the community health approach that stresses various variables that determine how long and how well we live. The data given allows us to assist communities in understanding how healthy citizens are and what type of health behaviors they possess.

### Motivation

Premature Mortality is a measure of length of life that focuses on preventable deaths, and this is done by increasing the weight of deaths of younger people. In the County Health Rankings data set, deaths that occur at ages significantly lower than 75 are counted more than deaths that occur at or above 75. This adjustment allows researchers to analyze which deaths are occurring because of life factors, such as poor health behaviors or physical environment, instead of old age. In the case of this analysis, it will allow focus to be on counties where citizens are exhibiting high rates of poor health behaviors but not on counties where the population is elderly.

Health behaviors are a leading cause of illness and death in the United states. Efforts to improve public health require information on the prevalence of health behaviors in populations. Studies conducted concluded that approximate half of all deaths in the Unites States could be attributed to factors such as smoking, physical inactivity, poor diet, and alcohol use. Identifying which health behaviors most contribute to premature deaths is is useful for health program planning and evaluation.

## Data

Our data set was gathered from the County Health Rankings Model. This model contains community health information that emphasizes the many factors that influence how long and how well we live. Specifically, we focused on county-level premature deaths and health behavior factors for our analysis.

Health behaviors are actions individuals take that affect their health. They include actions that lead to improved health, such as eating well and being physically active, and actions that increase one’s risk of disease, such as smoking, excessive alcohol intake, and risky sexual behavior. Below we provide descriptions of each variable within every health behavior category.

### Data Description

**Response**

- Premature Death: Years of potential life (YPLL) lost before age 75 per 100,000 population (2018-2020) 

**Predictors**
  
*Alcohol* 

- Excessive Drinking: Percentage of adults reporting binge or heavy drinking per 100,000 population (2019)
  
- Alcohol-Impaired Driving Deaths: Percentage of driving deaths with alcohol involvement (2016-2020)

*Diet and Exercise*

- Adult Obesity: Percentage of the adult population that reports a body mass index (BMI) greater than or equal to 30 kg/m2 (2019)
  
- Food Environment Index: Index of factors that contribute to a healthy food environment, from 0 (worst) to 10 (best)  (2019)
  
- Physical Inactivity: Percentage of adults age 18 and over reporting no leisure-time physical activity (age-adjusted) (2019)
  
- Access to Exercise Opportunities: Percentage of population with adequate access to locations for physical activity (2010 & 2021)
  
- Food Insecurity: Percentage of population who lack adequate access to food (2019)
  
- Limited Access to Healthy Foods: Percentage of population who are low-income and do not live close to a grocery store (2019)
  
*Tobacco Use*

- Adult Smoking: Percentage of adults (age 18 and older) who are current smokers (2019) 

*Sexual Activity* 

- Sexually Transmitted Infections (STI): Number of newly diagnosed chlamydia cases per 100,000 population (2019)

*Population*

- County Class: County description categorized by population (Urban $\geq$ 200k, Rural $\leq$ 50k, 200k $<$ Suburban $<$ 50k) (2020)

### Exploratory Data Analysis

We first started by exploring the geographical distribution of our response variable, YPLL, using choropleths. We developed a comprehensive function that outputs these YPLL maps reactively based on demographics and geographic region but for the purposes of the report we've only included the YPLL distribution across the entire United States.

```{r}
#YPLL Map
ypll_map
```

Next we modeled our predictors' relationships with the response variable. To do this comprehensively yet efficiently, we created pairs plots and a correlation matrix. For the purposes of this report, we've included only some scatter plots between our response and predictors that we believe captures the most significant relationships. Of our explanatory variables that demonstrated a significant relationship with YPLL, their relationship was largely linear with some curvature at the ends.

```{r}
#YPLL vs Predictors (use cowplot for formatting plots)
plot_grid(ypll_drink, ypll_smoke,ypll_food, ypll_obesity, 
          ncol = 2, labels = c('A', 'B', 'C', 'D'), label_fontface = 'bold',label_size = 11)
```

Given that county class is our only categorical predictor, we created a violin plot overlayed on top of a boxplot to model the relationship between county types and YPLL. Interestingly, we found that the mean YPLL and YPLL variability increases as population decreases. The mean increasing could be potentially due to poorer access to high quality healthcare in more rural areas versus the city or other factors. However, the high degree of variation in the YPLL could be explained by how the statistic is derived. YPLL measures years of potential lives lost per 100k people. Therefore in counties where their total population is 10k, their YPLL statistic will be multiplied by a factor of 10. Therefore, in smaller populations we are more likely to see high YPLL variability 'technically' at both ends (but from this plot we really only see that it is heavily skewed right).

```{r}
#YPLL vs Predictors colored by county class
ypll_countyclass #violin overlay on box plot, variability limitation in YPLL calculation for low pop counties
```

To test for collinearity we plotted the predictors against each other. We found that the diet/exercise variables were positively correlated with each other which could potentially complicate capturing inference from our models. We also saw that excessive drinking shared an inverse relationship with food insecurity. On the other hand, alcohol-impaired driving deaths and smoking were independent of each other as their scatterplot resembled a random scatter with zero slope.

```{r}
#Predictors vs Predictors (diet/diet)
plot_grid(food_obesity, food_drink, smoke_dui, 
          ncol = 2, labels = c('A', 'B', 'C'), label_fontface = 'bold', label_size = 11)
```

## Methods

Given the relatively linear relationships discovered in our EDA, we explored linear regression, general additive models (GAM), and regularization techniques. Additionally, to explore potential interactions between predictors we built a flexible tree based model using gradient boosted machines (GBM).

### Linear Regression

We designed a linear regression model using all of our predictors to serve as a baseline model for future model selection. Given the general linear relationships among YPLL and our predictors, we expect model to perform relatively well compared to more robust modeling techniques.

### Regularization Techniques

We explored Lasso, Ridge, and Elastic Net modeling to discourage learning more complex models and reduce the risk of overfitting. Given the collinearity between our exercise/diet related predictors, we designed Ridge and elastic net models to avoid potentially randomly removing highly collinear variables.


### General Additive Model

GAM are flexible models that are sure to outperform the linear regression because it could capture both linear and nonlinear relationships with its basis functions. 

Using backwards stepwise variable selection, I began with a saturated GAM using all predictors. I then removed the least significant predictor (based on p-value) and calculated the root mean squared error (RMSE) of this model. I continued this process until I exhausted all the predictors and I selected the model with the least RMSE.

This stepwise technique led us to create a simplified GAM with 7 features: Physical Inactivity, Obesity, Food Insecurity, Excessive Drinking, Smoking, STI, and County Class Descriptions.

### Gradient Boosted Machines

Add GBM to cv


## Results

We first explored the performance of our linear regression model using all 13 features compared to a lasso, ridge, and elastic net techniques. Using 5-fold stratified cross validation, we evaluated model performance on predicting YPLL using RMSE as our cost function. Using stratified cross validation ensures that relative county frequencies within group (i.e., state) is approximately preserved in each train and validation fold.

```{r}
#CV Plot
reg_cv
```

Despite the purpose of using regularization approaches being to reduce overfitting/variance, our linear regression model outperformed all three regularization techniques on the holdout data within intervals. This led us to explore a non-parametric approach that could still preserve linear relationships in the data using GAM.

We initially built a GAM on the same 13 predictors as the linear regression model. After performing feature selection, we also constructed a simplified GAM using only 7 of the most important features. We then performed 5-fold stratified cross validation to compare the performances of the linear regression, full GAM, simplified GAM and GBM using RMSE.

```{r}
#CV Plot
gam_cv
```

Both GAM models outperform the linear regression model within intervals. The simple GAM (7 features) outperforms the full GAM (11 features) within intervals. The simplified GAM's better performance is likely due to reduce model variability compared to the full GAM.

When plotting our fitted values on a holdout set of actual values for YPLL we found that are simple GAM model predicts well for a majority of the data when YPLL < 15,000. However, once YPLL > 15,000 our model tends to underestimate these observations. As stated earlier when evaluating YPLL versus county types, YPLL observations tend to artificially inflate for low population counties. This phenomena can explain why our model underestimates large YPLL values.

```{r}
#Fitted vs Actual
fin_mod
```

## Discussion

**Conclusions**

- Our simplified GAM outperforms our full GAM and linear regression model in predicting premature deaths.

- Our simplified GAM underestimates true YPLL values above 15,000 years.

**Limitations**

- The high variation in premature deaths among small populations (i.e., Rural counties) can create inflated and unreliable calculations for YPLL. This can lead to misguided conclusions in both the EDA and modeling process when mapping the relationship between the response and predictor.

- The years our response and predictor variables were collected vary. This complicates drawing direct inference of health behaviors relationship with premature deaths. For prediction purposes, we believe it is valid to use variables from varying years because time is not independent across counties. Therefore, health behaviors-- such as the proportion of tobacco smokers-- shouldn't significantly vary within a couple years within the same county.

- Due to anonymity constraints, we used county level health data rather than individual level data. However, data aggregated at the county level significantly underestimates variability in the data and does not give us insight into the demographic features of our predictor variables. With individual level data, we could group data by race and build models that can predict premature deaths by race using these grouped features.

**Future Work**

- Given the scope of this project, we focused on health behaviors' relationship with premature deaths. However, health behaviors only capture a piece of the factors that influence premature deaths. In the future, building a robust model using other health factors (e.g., clinical care, socioeconomic factors) to predict premature deaths could create a more complete picture.  

- In cohort studies, health data is collected from individuals among a specific area or demographic. Using a cohort study, the predictive model could show health behavior effects on premature mortality at the individual level and we could account for more specific health measures, such as blood pressure and waist circumference. 

## References
