---
title: "EDA Project"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This project will begin on Monday June 13th, and conclude with a 10-15 minute presentation on Friday, June 24th (either during the morning session from 10:30 to 12 PM or in the afternoon from 1:30 to 3 PM). The goal of this project is to practice understanding the structure of a dataset, and to practice generating and evaluating hypotheses using fundamental EDA and data visualization techniques.

Example Project: https://eryn184.github.io/cmsacamp-eda-project/NHL-Shots-EDA-Presentation.html#1

# Deliverables

Your team is expected to produce R Markdown slides (an example template will be provided shortly) to accompany your 10-15 minute presentation with the following information:

1. Explanation of the data structure of the dataset

2. Three hypotheses you are interested in exploring

3. Three data visualizations exploring the hypotheses, at least two of which must be multivariate. Each visualization must be in a different format from the other two, and you must have at least one categorical and one continuous visualization

4. One clustering example

5. Conclusions reached for the hypotheses based on your EDA and data visualizations

# Timeline

There will be two submission deadlines:

Friday, June 17th @ 5:00 PM EST - Each student will push their individual code for the project thus far to their GitHub accounts for review. We will then provide feedback on the code submitted.

Thursday, June 23rd @ 11:59 PM EST - Slides and full code must be completed and ready for presentation. Send your slides to Prof Yurko’s email (ryurko@andrew.cmu.edu). All code, visualizations, and presentations must be made in R. Take advantage of examples from lecture and the presentation template, but also feel free to explore material online that may be relevant!

# Data

Your team is assigned the Maternal Health Care Disparities data. The Centers for Disease Control and Prevention WONDER program helps track information related to birth records, parent demographics and risk factors, pregnancy history and pre-natal care characteristics. This data source could help identify combinations of risk factors more commonly associated with adverse outcomes which could then be utilized to develop better pre-natal care programs or targeted interventions to reduce disparities and improve patient outcomes across all ethnicities.

The data set is a sample of data from the CDC Wonder database for available birth records from 2019 that has been aggregated by state and a few conditions: (the number of prior births now deceased and whether or not the mother smoked or had pre-pregnancy diabetes or pre-pregnancy hypertension). So for example, the first row corresponds to the set of births that were born to women in Alabama who had no prior births deceased, smoked, and had both diabetes and hypertension pre-pregnancy. There were 12 such births, and the following variables (e.g. mother’s age) describe the mothers/infants in that set of 12.

State: State
PriorBirthsNowDeceased: The number of prior births now deceased
TobaccoUse: Whether or not the mother uses tobacco products
PrePregnancyDiabetes: Whether or not the mother had diabetes prior to becoming pregnant
PrePregnancyHypertension: Whether or not the mother had hypertension prior to becoming pregnant
Births: The number of births in that state with a defined combination of the previous four conditions (PriorBirthsNowDeceased, TobaccoUse, PrePregnancyDiabetes, PrePregnancyHypertension)
AverageMotherAge: The average mother’s age for the corresponding group of births
AverageBirthWeight: The average birth weight in grams for the corresponding group of births
AveragePrePregnancyBMI: The average pre-pregnancy BMI of the mother for the corresponding group of births
AverageNumberPrenatalVisits: The average number of prenatal visits of the mother for the corresponding group of births
AverageIntervalSinceLastBirth: The average length of time since the last birth for the corresponding group of births

```{r}
#read data
maternity_df = read.csv('maternity.csv')
```

# Project Start

```{r}
library(tidyverse)
```

## Dataset Descripition

```{r}
head(maternity_df)

#sum(maternity_df$Births) #3,744,389 individual obs

#dim(maternity_df) #842 obs, 13 var

#Do i have to list all var or only relevant var, how to display good looking data table in html
```

```
The Maternal Health Care Disparities Data

Loaded in a dataset of available domestic birth records from 2019

 - There's 3.7M observations aggregated into 842 rows

 - Rows represent unique state, prior births deceased, tobacco use, diabetes, and hypertension combinations

 - Categorical Variables: State, PriorBirthsNowDeceased, TobaccoUse, PrePregnancyDiabetes, PrePregnancyHypertension

 - Continuous Variables: Births, AverageMotherAge, AverageBirthWeight, AveragePrePregnancyBMI, AverageNumberPrenatalVisits,
                         AverageIntervalSinceLastBirth
```

## Hypothesis

1. Tobacco use reduces the child's birth weight among mother's without prior birth deaths

2. As the age of the mother increases, the number of prenatal visits increases.

3. Tobacco use is more prevalent among mothers with at least 1 prior birth deceased.

## Data Visualizations

```{r}
#Hypothesis 1

#create dataset with counterfactual
h1_df = filter(maternity_df, TobaccoUse %in% c("Yes", "No"), PrePregnancyDiabetes == "No", PrePregnancyHypertension == "No",
               PriorBirthsNowDeceased == "0")

#Calculate standard deviation in group using rep(AverageBirthWeight, births)
h1_df_no = filter(h1_df, TobaccoUse == "No")

no_hold = NULL

for(i in 1:length(h1_df_no)){
  
  temp = rep(h1_df_no[i,]$AverageBirthWeight, h1_df_no[i,]$Births)
  
  no_hold = c(no_hold, temp)  
}

h1_df_yes = filter(h1_df, TobaccoUse == "Yes")

yes_hold = NULL

for(i in 1:length(h1_df_yes)){
  
  temp = rep(h1_df_yes[i,]$AverageBirthWeight, h1_df_yes[i,]$Births)
  
  yes_hold = c(yes_hold, temp)
  
}

#Barplot with CI (The CI are so small bc the n is so large)
h1_df %>%
  group_by(TobaccoUse) %>%
  summarise(true_count = sum(Births), 
            mean_AverageBirthWeight = mean(AverageBirthWeight)) %>%
  mutate(total = sum(true_count),
         sd = c(sd(no_hold), sd(yes_hold)),
         margin = qt(0.975, df = true_count - 1)*sd/sqrt(true_count),
         lower = mean_AverageBirthWeight - margin,
         upper = mean_AverageBirthWeight + margin) %>%
  ggplot(aes(x = TobaccoUse, y = mean_AverageBirthWeight)) +
  geom_col(col = "black", fill = "steelblue") +
  #geom_errorbar(aes(ymin = lower,ymax = upper), color = "red") + #CI
  #coord_cartesian(ylim = c(3000, 3500)) + #Zoom in
  geom_text(aes(label = round(mean_AverageBirthWeight, 0)), vjust = -0.25) + #text label
  theme_bw() + 
  labs(title = "Tobacco Use is Associated with Decreased Birthweights", x = "Tobacco Use", y = "Average Birthweight (Grams)")
```

```
This barplot supports our hypothesis that tobacco use reduces the child's birth weight among mother's without prior birth deaths. Mother's with Tobacco use birth babies approximately 213 grams lighter than mother's without tobacco use. With 3.3. million observations for mother's not using tobacco and 207,000 observations for mother's using tobacco, there is sufficient sampling to state that this difference is statistically significant (the CI intervals are super small).
```

```{r}
#Hypothesis 2


```

```

```

```{r}
#Hypothesis 3
positions = c("None", "At least One")

filter(maternity_df, PriorBirthsNowDeceased != "Unknown", TobaccoUse != "Unknown") %>%
  mutate(new_dead_baby = ifelse(PriorBirthsNowDeceased == "0", "None", "At least One")) %>%
  group_by(new_dead_baby, TobaccoUse) %>%
  summarise(count = sum(Births)) %>%
  mutate(total = sum(count),
         prop = round(count/total, 5)) %>%
  ggplot(aes(x = new_dead_baby, y = prop, fill = TobaccoUse)) +
  geom_col() + 
  scale_x_discrete(limits = positions) + #set positions of bars
  scale_y_continuous(breaks = seq(0, 1, by = 0.2)) +
  theme_bw() +
  labs(title = "", x = "Number of Deceased Births", y = "Proportion of Births", fill = "Tobacco Use")
```

```
This visualization supports our hypothesis that tobacco use is more prevalent among mothers with at least 1 prior birth deceased. Among mother's with at least 1 prior birth deceased, approximately 6% used tobacco while only 1% of mother's without prior birth deaths used tobacco products.
```

## Clustering

```{r}
#Cluster based on two continuous variables (e.g., BMI, Birth Weight)
```

## Final Slide

**Conclusions**

H1

H2

H3

**Limitations**

We did not perform any statistical tests (e.g., t-test) to compare group averages, so our conclusions are purely based on visualizations.

Our clustering doesn't weigh the births within each row and rather treats each row as a single observation.

Feel free to ask any questions!

